Smart Spreading Factor Assignment for LoRaWANs

### 2019-01-13 paper.tex
omit lora details

\par There are several emerging LPWAN technologies. LoRa, Sigfox, NB-IoT and LTE-M are commonly used, well-known LPWAN technologies. LoRa and Sigfox use license free ISM frequency bands while NB-IoT and LTE-M use licensed frequency bands which brings extra cost \cite{7815384}. Both LoRa and Sigfox are known for ultra-low power consumption and resilience to interference.  While NB-IoT and LTE-M are promoted for higher data rate. LoRa has open standard MAC protocol called LoRaWAN. LoRaWAN and Sigfox MAC protocols are based on pure ALOHA medium access. LoRaWAN networks can be deployed as a private network like WiFi. However, Sigfox and NB-IoT are only available with operator contract \cite{7815384}. Number of messages that Sigfox end device can send in a day is limited to 140 packets for uplink and just 4 packets for downlink. Also, Sigfox packet payload size is limited to 12 bytes for uplink and 8 bytes for downlink. However, LoRa supports up to 243 bytes payload size and NB-IoT supports up to 1600 bytes payload size. Sigfox maximum data rate is 100 bps, on the other hand maximum data rate for LoRa and NB-IoT are 50 kbps and 200 kbps respectively \cite{7815384}.

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and transmission time on air \cite{7803607}. LoRa supports 6 different SF option. Simultaneous same SF transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other up to some extent, thus SF assignment is crucial for overall network performance. In this work, a custom made LoRa discrete event simulator is introduced to study different LoRa SF assignment strategies. We also proposed an SF assignment scheme to increase network performance with the help of machine learning methods and we verified our proposal using the simulator.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa is a wireless link technology for LPWAN applications. LoRa has an open standard MAC protocol called LoRaWAN. LoRa and LoRaWAN terms are frequently yet incorrectly used for each other.

\par LoRaWAN is a medium access control (MAC) layer protocol which designed for large scale LoRa networks. LoRaWAN is an open source standard developed and it is maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRa can be used as a wireless link technology without complying LoRaWAN, however LoRaWAN is developed considering well known LPWAN challenges and their best practice solutions. LoRaWAN also provides inter-operability between different LoRa networks. LoRaWAN provides lightweight but powerful standard for wide range of LoRa IoT applications.  LoRaWAN is based on pure ALOHA medium access which means end nodes do not check whether the channel is free or not before transmitting, accepting the possibility of the collision.

@@

\par In \cite{s17061193}, the authors investigated single gateway LoRaWAN network scalability in terms of the number of end nodes using a simulation model based on real measurements. They measure the impact of two concurrent LoRa transmissions on each other by using physical LoRaWAN end devices and a gateway then they created a simulation model from their measurements. Their results show that LoRaWAN has better scalability than pure ALOHA since a LoRa packet may still go through under collision if the last six symbols of preamble and header of the packet does not collide.

@@

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes SFs requires different rules. For this reason, we proposed a machine learning based SF assignment scheme to decrease same SF transmission collisions. This technique learns behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than the GW sensitivity. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the lowest possible SF to nodes considering the collision probability.

\par In this work, a decision tree classifier is used to predict transmission results. Gini index criteria is used for measuring quality of decision tree splits. As in most of the machine learning applications, increasing the data set size increases the prediction accuracy. For in this case, data set size is directly proportional to simulation duration. Thus, increasing simulation duration, increases the prediction accuracy. For real world cases, NS can keep track of transmission and it can create a classifier daily basis, then GWs can request from nodes to use this new SF.

\par We have been able to generate mass amount of LoRaWAN transmission logs for different topologies using the simulator. To come up with a solution for SF assignment issue, a machine learning prediction scheme is integrated to the simulator. A classifier is trained from transmission logs and this classifier is used for selecting best possible SF for the nodes. Simulator first runs a simulation with assigning random SF to nodes. Simulation results are combined into four data columns which are; X and Y coordinates of the transmission source, SF of the transmission and result of the transmission. Transmission result can be successful, interfered or under sensitivity. These data are fed to Python scikit-learn decision tree classifier. After the classifier model is generated, a second simulation is run with classifier. Simulator selects the lowest possible SF considering prediction of transmission. If a transmission of a node with lowest possible SF is predicted as interfered transmission by classifier, then simulator assigns a higher SF. If no SF is predicted as successful, then simulator again assigns lowest possible SF.

@@

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictates duty cycle transmission limit for all devices including GWs. This makes downlink transmissions to rare.

@@

\begin{table}
\centering
\caption{EU863-870 ISM Band LoRa Default Channels \cite{lorawan.regional.parameters}.}
\label{table:max_tx_power}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Channel} & \textbf{Frequency (MHz)} & \textbf{Bandwidth (kHz)} & \textbf{Max EIRP (dBm)} \\ \hline
      1 &     868.1 &   125 &   14 \\ \hline
      2 &     868.3 &   125 &   14 \\ \hline
      3 &     868.5 &   125 &   14 \\

\hline
\end{tabular}
\end{table}

\par Maximum transmit power for European ISM band LoRa default channels can be found in Table \ref{table:max_tx_power}. In the simulator, it is assumed that nodes always select maximum allowed transmit power which is 14 dBm. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in the simulator.

\begin{table}
\centering
\caption{Gateway Sensitivity for Different SFs \cite{SX1276}.}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{}} & \multicolumn{6}{c|}{\textbf{SF}} \\ \cline{3-8}
\multicolumn{2}{|c|}{}                  &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
\multirow{3}{*}{\textbf{BW (kHz)}}  & 125 & -123 & -126 & -129 & -132 & -133 & -136 \\ \cline{2-8}
                                    & 250 & -120 & -123 & -125 & -128 & -130 & -133 \\ \cline{2-8}
                                    & 500 & -116 & -119 & -122 & -125 & -128 & -130 \\ \hline
\end{tabular}
\end{table}

\par Receive sensitivity of a LoRa gateway for different SFs and BWs in dBm unit can be found in Table \ref{table:gw_sf_sensitivity} \cite{SX1276}. In the simulator, 125 kHz bandwidth is used which is allowed bandwidth in EU863-870 regulation.

\par Free space propagation loss can be calculated as \cite{TR136.942}:

\begin{equation} \label{eq:propagation_loss}
\begin{split}
P^{dB}_{PATH} = 40(1 - 4 \times 10^{-3} \times h){\log_{10} R|_{km}} \\
- 18 {\log_{10} h|_{m}} + 21 {\log_{10} f|_{MHz}} + 80
\end{split}
\end{equation}

\par Where, $h$ is the gateway altitude and $f$ is the frequency of the signal. In this work, it is assumed that $h$ = 15 m and $f$ = 868 MHz. With these assumptions, propagation loss calculation become \cite{7996384}:

\begin{equation} \label{eq:propagation_loss_simplified}
P^{dB}_{PATH} = 120.5 + 37.6 {\log_{10} R|_{km}}
\end{equation}

@@

\subsection{Interference Model}
\par In the simulator, interference model described in \cite{7996384} is adopted. They introduced SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions.

\par In our simulator, it is assumed that there is no other technology interference is present in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, following signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used:

\begin{equation} \label{eq:sinr}
T = \begin{bmatrix}
       6 & -16 & -18 & -19 & -19 & -20 \\
     -24 &   6 & -20 & -22 & -22 & -22 \\
     -27 & -27 &   6 & -23 & -25 & -25 \\
     -30 & -30 & -30 &   6 & -26 & -28 \\
     -33 & -33 & -33 & -33 &   6 & -29 \\
     -36 & -36 & -36 & -36 & -36 &   6
     \end{bmatrix}
\end{equation}

\par To decide if a referenced signal is interfered at receiver by a interferer signal, SINR threshold matrix is used. $T_{i,j}$ is the needed SINR margin in dB unit between referenced signal with SF = i and interferer signal with SF = j to correctly decode the referenced signal. If there are more than one interferer, referenced signal must satisfy the margin for cumulative sum of all interfering signal received powers for each SF \cite{7996384}. SINR threshold can be calculated as:

\begin{equation} \label{eq:sinr_db}
SINR_{i,j} = \dfrac{P_{rc,0}}{\sum_{l \in I_j} P_{rc,l}}
\end{equation}

\par Where $P_{rc,0}$ is received signal power of referenced signal and $P_{rc,l}$ is received signal power of interfering signal at SF = j. If a packet with SF = i satisfies the following condition for every SF = j, then packet is survived from all interference.

\begin{equation} \label{eq:sinr_t}
SINR_{i,j}^{dB} > T_{i,j}
\end{equation}

\par To calculate interfering power at receiver, we should consider the case where two transmissions are not perfectly overlapping. To equalize the interfering power \cite{7996384}:

\begin{equation} \label{eq:p_interference}
P_{rc,y}^{interf} = \dfrac{P_{rc,y}(t_{interf})}{t_{x}}
\end{equation}

\par Where $t_{x}$ is transmission duration of referenced signal. $t_{interf}$ is intersection duration between referenced signal and interfering signal. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}. Duration of a LoRa transmission can be calculated as:

\begin{equation} \label{eq:transmission_duration}
t_{x} = \dfrac{8 \times PS|_{B}}{R_{b}|_{bps}}  \ s
\end{equation}

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows. Packet size is 60 bytes including header and payload. Simulation duration is 3600 seconds.

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters, number of GW is set to 1 and packet generation rate is set to 0.01 packet per second. Increasing SFs increases time on air. This increases the number of collisions thus decreases the PDR of the network. PDR of higher SFs are decreased to almost 0 while number of nodes is increasing. Since network topology radius is quite small almost all SFs can reach to the GW.

@@

\par In Figure \ref{fig:gw_pdr}, PDR plot of different number of GW is shown. Randomly generated topology radius is set to 3000 meters, packet generation rate is set to 0.01 packet per second and lowest possible SF assignment scheme is used. Increasing number of GWs, decreases the SFs of nodes and decreases time on air. This decreases number of collisions thus increases the PDR of the network when network topology radius is constant.

@@

\par In Figure \ref{fig:r_pdr}, PDR plot of different network topology radius is shown. Number of GW is set to 1, packet generation rate is set to 0.01 packet per second and lowest possible SF assignment scheme is used. Increasing network topology radius, increases the number of under sensitivity transmissions thus decreases the PDR of the network.

@@

\par In Figure \ref{fig:pr_pdr}, PDR plot of different packet generation rate is shown. Randomly generated topology radius is set to 3000 meters, number of GW is set to 1 and lowest possible SF assignment scheme is used. Increasing packet generation rate, increases the number of collisions thus decreases the PDR of the network.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{{pr_pdr_r3000_g1_s3600}.png}
\caption{PDR of different packet rates.}
\label{fig:pr_pdr}
\end{figure}

\par In Figure \ref{fig:prediction_pdr}, PDR plot of machine learning prediction scheme is shown. Randomly generated network topology radius is set to 5000 meters, number of GW is set to 3 and packet generation rate is set to 0.01 packet per second. Prediction model needs nodes location so three GWs are used for prediction SF scheme since three GWs are enough to roughly locate nodes positions by triangulation. Prediction scheme gives better PDR when number of nodes increases since number of interferences also increases. Prediction scheme can only improve network performance when LoRa interference is present. Also, prediction scheme gives better results when nodes are deployed close to the gateway, since nodes have margin to increase their SFs. If a node is far away from the GW, then prediction scheme cannot increase the SF to avoid interference since the SF is already high.

@@

\begin{table}
\centering
\caption{PDR of lowest SF selection and different SF classifier schemes.}
\label{table:prediction_pdr}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{\multirow{2}{*}{}}                            & \multicolumn{3}{c|}{\textbf{Number of Nodes}} \\ \cline{4-6}
\multicolumn{3}{|c|}{}                                             & 100           & 500           & 1000          \\ \hline
\multirow{12}{*}{\textbf{R (m)}} & \multirow{3}{*}{3000}  & Lowest & 97.8          & 86.0          & 72.3          \\ \cline{3-6}
                                 &                        & SVM    & 98.0          & 88.2          & 75.2          \\ \cline{3-6}
                                 &                        & DTC    & 97.8          & 89.8          & 78.7          \\ \cline{2-6}

@@

\begin{table}
\centering
\caption{Prediction accuracy (\%) of different classifiers.}
\label{table:prediction_accuracy}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{\multirow{2}{*}{}}                        & \multicolumn{3}{c|}{\textbf{Number of Nodes}} \\ \cline{4-6}
\multicolumn{3}{|c|}{}                                         & 100           & 500           & 1000          \\ \hline
\multirow{8}{*}{\textbf{R (m)}} & \multirow{2}{*}{3000}  & SVM & 82.4          & 70.4          & 71.7          \\ \cline{3-6}
                                &                        & DTC & 86.0          & 67.3          & 70.4          \\ \cline{2-6}

@@

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume that nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmissions to more vulnerable to interference thus requires extra care. Also, other machine learning methods can be investigated for SF optimization. Reinforcement learning is a good candidate for this network optimization issue. TODO, node id without location TODO

>>>>

\par There are several emerging LPWAN technologies. LoRa, Sigfox, NB-IoT and LTE-M are commonly used, well-known LPWAN technologies. LoRa and Sigfox use license free ISM frequency bands while NB-IoT and LTE-M use licensed frequency bands which brings extra cost \cite{7815384}. Both LoRa and Sigfox are known for ultra-low power consumption and resilience to interference. While NB-IoT and LTE-M are promoted for higher data rate. LoRa has open standard MAC protocol called LoRaWAN. LoRaWAN and Sigfox MAC protocols are based on pure ALOHA medium access. LoRaWAN networks can be deployed as a private network like WiFi. However, Sigfox and NB-IoT are only available with operator contract \cite{7815384}.

\par LoRa can adjust data rate by spreading symbols within a fixed channel bandwidth. This enables tradeoff between receive sensitivity and transmission time on air \cite{7803607}. Simultaneous same spreading factor (SF) transmissions are prone to collision, however, different SF transmissions in the same channel are orthogonal to each other, thus SF assignment is crucial for overall network performance. In this work, a LoRa discrete event simulator is developed to study different LoRa SF assignment strategies. We also proposed an SF assignment scheme to increase network performance with the help of machine learning methods and we verified our proposal using the simulator.

@@

\section{LoRaWAN} \label{LoRaWAN}
\par LoRa has an open standard medium access control (MAC) layer protocol called LoRaWAN. LoRaWAN is a medium access control (MAC) layer protocol which designed for large scale LoRa networks. LoRaWAN is developed and maintained by LoRa Alliance. LoRa Alliance is an open, non-profit organization dedicated to standardization of LoRaWAN. LoRaWAN  provides inter-operability between different LoRa networks. LoRaWAN is based on pure ALOHA medium access which means end nodes do not check whether the channel is free or not before transmitting, accepting the possibility of the collision.

@@

@@

\par It is difficult to propose a single SF assignment rule for every possible LoRaWAN topology since every network is different and optimizing their nodes' SFs requires different rules. For this reason, we proposed a machine learning based SF assignment scheme to decrease same SF transmission collisions. This technique works by learning the behavior of the nodes in a network. A NS can keep track of successful uplink transmissions and their SFs. NS can also keep track of some of the collided transmission if header part of the packet is not interfered at location of the GW. However, NS cannot keep track of transmissions with receive power lower than the GW sensitivity. Using these obtained information, NS can train a classifier for predicting future transmission result for specific node and specific SF. Using this prediction model NS can assign the SF to nodes considering the collision probability.

\par In this work, decision tree classifier (DTC) and support vector machine (SVM) are evaluated to predict transmission results. Gini index criteria is used for DTC splits. Class weights are balanced according to sample distributions for both methods. Data set size is directly proportional to simulation duration. Thus, increasing simulation duration, improves the prediction accuracy. For real world cases, NS can keep track of transmission and it can create a classifier daily basis, then GWs can request from nodes to use their new SFs.

\par We have been able to generate mass amount of LoRaWAN transmission logs for different topologies using the simulator. To come up with a solution for SF assignment issue, DTC and SCM prediction schemes are integrated to the simulator. A classifier is trained from transmission logs and this classifier is used for selecting best possible SF for the nodes. Simulator first runs a simulation with assigning random SF to the nodes. Simulation results are combined into four data columns which are; X and Y coordinates of the transmission source, SF of the transmission and result of the transmission. Transmission result can be successful, interfered or under sensitivity. These data are fed to Python scikit-learn DTC or SVM classifiers. After the classifier model is generated, a second simulation is run with classifier. Simulator selects the most suitable SF considering prediction of transmission. If a transmission of a node with lowest possible SF is predicted as interfered by the classifier, then simulator assigns a higher SF. If no SF is predicted as successful, then simulator again assigns the lowest possible SF.

TODOOOOO

@@

\par In the simulator, only LoRaWAN Class A devices are covered. Transmissions are always initiated by end nodes in pure ALOHA manner. Nodes generate a new packet according to Poisson interval for given packet rate input parameter. In the simulator, downlink transmissions are not considered. Downlink transmissions are rare in real world deployments since ISM band regulations dictates duty cycle limit for all devices including GWs.

@@

\par In the simulator, it is assumed that nodes always select maximum allowed transmit power, which is 14 dBm for European ISM band regulation. Different channel transmissions are independent from each other. But, in this work we focus on SF orthogonality. Thus, we only utilized single channel transmissions in the simulator.

\par Receive sensitivity of a LoRa gateway for different SFs and BWs in dBm unit can be found in \cite{SX1276}. In the simulator, 125 kHz bandwidth receive sensitivities are used.

\par Free space propagation loss is calculated as in \cite{TR136.942}. In this work, it is assumed that $h$ = 15 m and $f$ = 868 MHz.

@@

\subsection{Interference Model}
\par In the simulator, interference model described in \cite{7996384} is adopted. They use SINR threshold matrix for modeling LoRa interference between simultaneous but different SF LoRa transmissions.

\par In our simulator, it is assumed that there is no other technology interference in the network except LoRa interference. To exploit imperfect orthogonality of different SF transmission, simulator should calculate the effect of different SF transmissions to each other. In simulator, following signal to interference plus noise ratio (SINR) threshold matrix from \cite{goursaud:hal-01231221} is used.

\par To decide if a referenced signal is interfered at receiver by a interferer signal, SINR threshold matrix is used. $T_{i,j}$ is the needed SINR margin in dB unit between referenced signal with SF = i and interferer signal with SF = j to correctly decode the referenced signal. If there are more than one interferer, referenced signal must satisfy the margin for cumulative sum of all interfering signal received powers for each SF \cite{7996384}. Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}.

\section{Simulation Results}  \label{Simulation Results}
\par For simulation results in this paper, global simulation parameters are set as follows. Packet size is 60 bytes including header and payload, simulation duration is 3600 seconds, packet generation rate is 0.01 packet per second.

\par In Figure \ref{fig:sf_pdr}, PDR plot of different SF assignment is shown. Randomly generated network topology radius is set to 3000 meters and number of GW is set to 1. Increasing SFs increases time on air. This increases the number of collisions thus decreases the PDR of the network. PDR of higher SFs are decreased to almost 0 while number of nodes is increasing. Since network topology radius is quite small almost all SFs can reach to the GW.

@@

\par In Figure \ref{fig:gw_pdr}, PDR plot of different number of GW is shown. Randomly generated topology radius is set to 3000 meters and lowest possible SF assignment scheme is used. Increasing number of GWs, decreases the SFs of nodes and decreases time on air. This decreases number of collisions thus increases the PDR of the network when network topology radius is constant.

@@

\par In Figure \ref{fig:r_pdr}, PDR plot of different network topology radius is shown. Number of GW is set to 1 and lowest possible SF assignment scheme is used. Increasing network topology radius, increases the number of under sensitivity transmissions thus decreases the PDR of the network.

@@

\par For space constraints, we omit figures for network throughput output, total transmit energy consumption output and different packet generation rate input. We invite readers to experiment the simulation tool with different simulation inputs and outputs. \cite{simlorasf}.

\par In Figure \ref{fig:prediction_pdr}, PDR plot of machine learning prediction scheme is shown. Randomly generated network topology radius is set to 5000 meters and number of GW is set to 3. Prediction model needs nodes location so three GWs are used for prediction SF scheme since three GWs are enough to roughly locate nodes positions by triangulation. Prediction scheme gives better PDR when number of nodes increases since number of interferences also increases. Prediction scheme can only improve network performance when LoRa interference is present. Also, prediction scheme gives better results when nodes are deployed close to the gateway, since nodes have margin to increase their SFs. If a node is far away from the GW, then prediction scheme cannot increase the SF to avoid interference since the SF is already high.

@@

\par In Table \ref{table:prediction_pdr}, TODO

\par In Table \ref{table:prediction_accuracy}, TODO

\begin{table}
\centering
\caption{PDR of lowest possible SF and smart SF assignment schemes.}
\label{table:prediction_pdr}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{\multirow{2}{*}{}}                            & \multicolumn{3}{c|}{\textbf{Number of Nodes}} \\ \cline{4-6}
\multicolumn{3}{|c|}{}                                             & 100           & 500           & 1000          \\ \hline
\multirow{12}{*}{\textbf{R (m)}} & \multirow{3}{*}{3000}  & Lowest & 97.8          & 86.0          & 72.3          \\ \cline{3-6}
                                 &                        & SVM    & 98.0          & 88.2          & 75.2          \\ \cline{3-6}
                                 &                        & DTC    & 97.8          & 89.8          & 78.7          \\ \cline{2-6}

@@

\begin{table}
\centering
\caption{Prediction accuracy (\%) of SVM and DTC classifiers.}
\label{table:prediction_accuracy}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{3}{|c|}{\multirow{2}{*}{}}                        & \multicolumn{3}{c|}{\textbf{Number of Nodes}} \\ \cline{4-6}
\multicolumn{3}{|c|}{}                                         & 100           & 500           & 1000          \\ \hline
\multirow{8}{*}{\textbf{R (m)}} & \multirow{2}{*}{3000}  & SVM & 82.4          & 70.4          & 71.7          \\ \cline{3-6}
                                &                        & DTC & 86.0          & 67.3          & 70.4          \\ \cline{2-6}

@@

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume that nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmissions to more vulnerable to interference thus requires extra care. Also, other machine learning methods can be investigated for SF optimization. Reinforcement learning is a good candidate for this network optimization issue. Other transmission parameters such as node id and transmission time can be included to proposed scheme to improve performance.
